### 2025-06-11 paper/relatedwork.tex
Reference verification paper.

While in particular the dedicated frameworks tend to find widespread 
acceptance by practitioners, both, correctness and performance leave
two open questions: (i) how is it possible to
ensure that the AD algorithm is implemented correctly?
(ii) if the
language does not perform as expected, what are the
options to solve this?  Unfortunately, in many cases the answer to
both questions is unsatisfying.  Most languages do not
come with formal correctness guarantees, so one has to trust the
implementers of these tools.  One can run tests as well to gain trust 
in the implementation but that is far from a 
formal guarantee.  If one relies
on the AD provided by a chosen language/framework, and the generated code does not
perform well, one has to modify the language/framework, as these solutions
are tightly integrated with the tools. The problem here is that most of of these tools
have very large and sophisticated implementations typically comprising
of hundreds of thousands of lines of code.  Note that while our work
does not give a formal proof that AD implementation is correct (it is
not clear how to formulate correctness criteria for it), the entire
implementation of AD is about 20 lines of code, so it is easy to
understand and modify.

>>>>

While in particular the dedicated frameworks tend to find widespread acceptance
by practitioners, both, correctness and performance leave two open questions:
(i) how is it possible to ensure that the AD algorithm is implemented correctly?
(ii) if the language does not perform as expected, what are the options to solve
this? Unfortunately, in many cases the answer to both questions is unsatisfying.
Most languages do not come with formal correctness guarantees, so one has to
trust the implementers of these tools. One can run tests as well to gain trust
in the implementation but that is far from a formal guarantee. If one relies on
the AD provided by a chosen language/framework, and the generated code does not
perform well, one has to modify the language/framework, as these solutions are
tightly integrated with the tools. The problem here is that most of of these
tools have very large and sophisticated implementations typically comprising of
hundreds of thousands of lines of code. Note that while our work does not give a
formal proof that AD implementation is correct, the entire implementation of AD
is about 20 lines of code, so it is easy to understand and modify. Formulating
correctness criteria for AD algorithms is somewhat involved, but has been done
for implementations where the AD algorithm monitors the program when it runs and
records a trace of its actions~\cite{de2023verifying}, while our approach is
based on static program transformation---this is in fact the only viable
approach when also doing program extraction.
